The role of cognateness in native spoken word recognition

### 2024-10-08 manuscript/manuscript.tex
Add discussion and data about knowledge filter

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{tipa}
\usepackage{booktabs}
\DeclareTextFontCommand\textphonetic{\ipafont}
\usepackage[utf8]{inputenc}
\usepackage{CJKutf8}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
\ifPDFTeX
\else
\babelfont{rm}[]{Latin Modern Roman}
\fi
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

When some German speakers listened to the song \emph{The Power} by SNAP!
(\emph{Coyote Ugly}, 2000), many of them misheard the line ``I've got
the power'' as \emph{Agatha Bauer!} When Michael Jackson vocalised
``ma-ma-coo-sah'' in his song \emph{Wanna Be Startin' Somethin'}
(\emph{Thriller}, 1982), some Dutch speakers understood \emph{mama
appelsap} {[}mama applejuice{]}. When Spanish speakers listen to the
line ``Circumvent your thick ego'' from the song \emph{Pictures} by
System of a Down (\emph{Steal this Album!}, 2002), they tend to hear
\emph{Sácame de aquí, King-Kong} {[}take me out of here, King-Kong{]}.
When Japanese speakers listen to the line ``I want to hold your hand''
in the homonym song by The Beatles (\emph{Meet The Beatles}, 1964), they
often mishear \emph{Aho na hōnyōhan} {[}stupid urinator{]}. Outrageous
as these examples might sound to speakers of other languages, many
readers may know a few cases in their own native language. This auditory
illusion can feel quite real, and often inevitable after the first time
it happens. In Japanese, this phenomenon takes the name \emph{Soramimi}
(lit. ``empty ear''), after the \emph{Soramimi Hour}---a section of the
program \emph{Tamori Club} hosted by comedian Morita Kazuyoshi
(Tamori)---in which instances of this illusion were presented with
comedy purposes. Soramimi is a particular case of \emph{homophonic
translation}: words or phrases in one language are translated into
similar-sounding words and phrases in a different language, without
necessarily preserving the meaning. This phenomenon extends beyond song
lyrics, and may occur anytime when listening to non-native speech. For
instance, when listening to ``You know me?'' (\textipa{/ju: n@U mi:}/)
in English, Japanese-native listeners tend to understand \emph{yunomi}
(/\textipa{jUnomi}/), which is a type of Japanese teacup. Overall, the
occurence of Soramimi---and more generally, homophonic
translations---suggests that speech in an unfamiliar language has the
potential to activate lexical representations in the native language.

Although homophonic translations do not necessarily take place with the
preservation of meaning---which often results in outrageous
mistranslations like the ones aforementioned---homophonic translations
sometimes may lead to the activation of the intended meaning (i.e., a
correct translation). For example, imagine a Spanish native with no
previous familiarity with any other language listening to French for the
first time. This listener may encounter the word /\textipa{"pOKt}/
(\emph{porte}), translation of door in French. When comparing the
phonemic transcription of the French to that of its Spanish translation
/\textipa{"pweR.ta}/ (\emph{puerta}), one may be led to infer that these
two words do not have much overlap. But one thing to note is that the
voiced uvular fricative consonant /\textipa{K}/ and the open-mid back
rounded vowel /\textipa{O}/, which do not exist in Spanish as phonemic
categories, can be perceived as allophonic variations the native
(Spanish) phonemes /\textipa{R}/ and /o/, as described by assimilation
models of phoneme perception (\citeproc{ref-best1988examination}{Best et
al., 1988}, \citeproc{ref-best2001discrimination}{2001}). Although this
phonetic mismatch between the non-native word and its translation can
have a noticeable toll on comprehension Cutler et al.
(\citeproc{ref-cutler2004patterns}{2004}), the Spanish listener may
still be able to activate the phonological representation of
/\textipa{"pweR.ta}/ (\emph{puerta}) from hearing the word-form
/\textipa{"pOKt}/ (\emph{porte}), ultimately allowing them to access the
correct semantic representation. Given enough phonological similarity
between the non-native word and its translation in the native language,
a naïve listener may succeed at word comprehension when listening to an
unfamiliar language. Such phonological similarity between translation
equivalents---known as \emph{cognateness}---is common across many
languages, and is often due to typological closeness and/or
socio-historical events involving the speakers of these languages (e.g.,
migration, social contact). For example, Romance languages such as
Spanish and Catalan share many form-similar translation equivalents
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and orthographic similarity of the former to the equivalent
in their native language.

While Soramimi, and more generally homophonic translation, are popular
and frequent phenomena, the psycholinguistic literature on the topic is
relatively scarce, and has mostly focused on the phonological variables
involved in the occurence of Soramimi. For instance, Otake
(\citeproc{ref-otake2007interlingual}{2007}) analysed 194 instances of
Soramimi, broadcasted between 1992 and 2007 by the TV show Soramimi
hour. These instances consisted of homophonic translations of English
song lyrics to Japanese words and phrases. The vast majority of the
reported instances were phrases (96\%) and a few, to single words (4\%).
The phonetic features of the presented English lyrics were preserved
with relatively high variability, with some Japanese resulting words or
phrases sharing high overlap with the original English strings of
sounds, and some sharing very little overlap. When analysing the few
instances of homophonic translations of single words, the author
identified three phonological processes that might explain how Japanese
listeners reconstructed the English input to generate the Japanese
words: insertion (e.g., \emph{cry} /\textipa{kraI}/ to \emph{kurai},
/\textipa{kuRai}/ {[}dark{]}), deletion (e.g., \emph{go},
/\textipa{goU}/ to \emph{go}, /go/ {[}go{]}, and alternation (e.g.,
\emph{low}, /\textipa{loU}/ to \emph{rou}, /\textipa{Roo}/ {[}wax{]}).
This suggests that the reported homophonic translations can be
explained, to some extent, as the result of the Japanese listeners
accommodating the strings of English sounds to the Japanese phoneme
inventory and Japanese phonotactics
(\citeproc{ref-dupoux1999epenthetic}{Dupoux et al., 1999};
\citeproc{ref-peperkamp2008perceptual}{Peperkamp et al., 2008}). Native
prosody also constrains the occurence of homophonic translations. For
instance, Kentner (\citeproc{ref-kentner2015rhythmic}{2015}) presented
in the auditory modality French-native and German-native participants
with English songs with lyrics. Homophonic translations followed
language-specific segmentation strategies. Most homophonic translations
reported by German-native participants resulted from inserting phrase
and word boundaries \emph{before} stressed syllables (following the more
frequent trochaic patter in German). On the contrary, most homophonic
translations reported by French-native participants resulted from
inserting phrase and word boundaries \emph{after} stressed syllables,
following the more frequent iambic pattern in French.

While the previous studies have address the phonological aspects of
homophonic translation, little attention has been paid to the dynamics
of lexical activation and competition that underlay this phenomenon. In
the present work we investigated the interplay between phonological
similarity and phonological neighbourhood density to examine how
listeners of an unfamiliar language activate lexical representations in
their native language. In particular, we studied how the presence of
phonological neighbours in the native language lexicon affects the
extent to which listeners are able to activate the correct translation
of an auditorily presented word in an unfamiliar language. We used a
translation elicitation task in which participants listened to words
from an unfamiliar language, and then tried to guess their translation
in their native language. We will henceforth refer to the
auditory-presented words heard by participants on each trial as
presented words, and the correct translation for the presented words as
target translations. We explored listeners' reliance on phonological
similarity to succeed in the task by manipulating the amount of
phonological similarity between the presented words and their target
translations. Since participants were unfamiliar with the presented
language, they could only use phonological similarity between the
presented language and their native language to successfully translate
the words. We therefore predicted that participants' performance should
increase when the translation pairs are phonologically more similar. We
also predicted that there is a minimum threshold of phonological
similarity to be sufficient for translation.

@@

In Experiment 1, we collected data from two groups of British English
natives living in the United Kingdom. One group was presented with
Catalan words, the other with Spanish words. We examined the extent to
which participants were able to use the phonological similarity between
the presented word (in Catalan or Spanish) and its target translation to
provide accurate responses. In Experiment 2, we tested a group of
(European) Spanish natives in the same task, who were presented with
Catalan words. Catalan and Spanish are both Romance languages whose
close typological distance is reflected in the fact that they share many
cognates, where English is a Germanic language that shares considerably
fewer cognates with Catalan and Spanish. By testing participants
translating words from typologically close or distant languages, we
expected to widen the range of the phonological similarity scores of the
translation pairs involved in the experimental task, therefore allowing
us to explore potential cross-language differences in participant's
performance. One unexpected finding was that participants in Experiments
1 and 2 were surprisingly good at translating a subset of words which
had low phonological similarity with their correct translation. We were
concerned that this may be caused by some prior knowledge of specific
words by our participants, as some Spanish words are commonly seen in
media or product labels, making them familiar even to monolingual
speakers of English. In Experiment 3, we collected additional data from
a new group of British English natives. The design was closely modelled
after Experiment 1, except that after providing their response in each
trial, participants reported whether they had previous knowledge of the
presented word. In the final section of this paper, we present analyses
on the joint data sets of Experiments 1 to 3.

@@

Participants were recruited via Prolific (5£ compensation) and SONA
(compensation in academic credits), and gave informed consent before
providing any data and the study was conducted in accordance with
ethical standards of the Declaration of Helsinki and the protocol was by
the University of Oxford Medical Sciences Inter-Divisional Research
Ethics Committee (IDREC) (R60939/RE007). Participants were asked to
complete the experiment using a laptop in a quiet place with good
internet connection. Data collection took place from June 04th, 2020 to
June 25th, 2020. We collected data from 71 British English native
participants living in United Kingdom (\emph{Mean} = 21.76 years,
\emph{SD} = 2.15, Range = 18-26, 46, 25 female).

@@

\caption{\label{tbl-participants}}

@@

We arranged two lists of input words to be presented to participants in
the auditory modality: one in Catalan and one in Spanish. Words in the
Catalan list were 5.02 phonemes long on average (\emph{SD} = 1.49, Range
= 2-8), and the orthographic forms of their English translations (which
participants had to type) were 5.12 characters long on average
(\emph{SD} = 1.56, Range = 3-9. Words in the Spanish list were 5.52
phonemes long on average (\emph{SD} = 1.47, Range = 3-9), and the
orthographic form of their English translations (which participants had
to type), were 5.29 characters long on average (\emph{SD} = 1.77, Range
= 3-12.

Participants listened to one audio file in each trial, each containing a
single word presented in isolation. The audio files were the same ones
used in child experiments conducted in the Laboratori de Recerca en
Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio
files were recorded by a proficient Catalan-Spanish female bilingual
from the Metropolitan Area of Barcelona in a child-directed manner.
Catalan and Spanish words were recorded at 44,100 Hz in separate files
in the same session, and then de-noised using Audacity and normalised at
peak intensity using Praat (\citeproc{ref-broersma2021praat}{Broersma \&
Weenink, 2021}). The average duration of the Catalan audio files was
1.24 (\emph{SD} = 0.19, Range = 0.8-1.58). The average duration of the
Catalan audio files was 1.16 (\emph{SD} = 0.15, Range = 0.78-1.53).

For each translation pair, we defined three predictors of interest: the
lexical frequency of the correct translation (\emph{Frequency}), the
phonological similarity between the presented (Catalan or Spanish) word
and their correct English translation (\emph{Similarity}), and the
presented word's number of cross-language phonological neighbours
(\emph{CLPN}).

We included \emph{Frequency} as a nuance predictor, under the hypothesis
that---keeping other predictors constant---participants would translate
higher-frequency words more accurately and faster than lower-frequency
words. Lexical frequencies of correct translations were extracted from
SUBTLEX-UK (\citeproc{ref-van2014subtlex}{Van Heuven et al., 2014}), and
transformed to Zipf scores. Words in the stimuli list without a lexical
frequency value were excluded from data analysis (2 in the Catalan list,
5 in the Spanish list).

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{tipa}
\usepackage{booktabs}
\DeclareTextFontCommand\textphonetic{\ipafont}
\usepackage[utf8]{inputenc}
\usepackage{CJKutf8}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
\ifPDFTeX
\else
\babelfont{rm}[]{Latin Modern Roman}
\fi
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage[english]{selnolig} % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

When some German speakers listened to the song \emph{The Power} by SNAP!
(\emph{Coyote Ugly}, 2000), many of them misheard the line ``I've got
the power'' as \emph{Agatha Bauer!} When Michael Jackson vocalised
``ma-ma-coo-sah'' in his song \emph{Wanna Be Startin' Somethin'}
(\emph{Thriller}, 1982), some Dutch speakers understood \emph{mama
appelsap} {[}mama applejuice{]}. When Spanish speakers listen to the
line ``Circumvent your thick ego'' from the song \emph{Pictures} by
System of a Down (\emph{Steal this Album!}, 2002), they tend to hear
\emph{Sácame de aquí, King-Kong} {[}take me out of here, King-Kong{]}.
When Japanese speakers listen to the line ``I want to hold your hand''
in the homonym song by The Beatles (\emph{Meet The Beatles}, 1964), they
often mishear \emph{Aho na hōnyōhan} {[}stupid urinator{]}. Outrageous
as these examples might sound to speakers of other languages, many
readers may know a few cases in their own native language. This auditory
illusion can feel quite real, and often inevitable after the first time
it happens. In Japanese, this phenomenon takes the name \emph{Soramimi}
(lit. ``empty ear''), after the \emph{Soramimi Hour}---a section of the
program \emph{Tamori Club} hosted by comedian Morita Kazuyoshi
(Tamori)---in which instances of this illusion were presented with
comedy purposes. Soramimi is a particular case of \emph{homophonic
translation}: words or phrases in one language are translated into
similar-sounding words and phrases in a different language, without
necessarily preserving the meaning. This phenomenon extends beyond song
lyrics, and may occur anytime when listening to non-native speech. For
instance, when listening to ``You know me?'' (\textipa{/ju: n@U mi:}/)
in English, Japanese-native listeners tend to understand \emph{yunomi}
(/\textipa{jUnomi}/), which is a type of Japanese teacup. Overall, the
occurrence of Soramimi---and more generally, homophonic
translations---suggests that speech in an unfamiliar language has the
potential to activate lexical representations in the native language.

Although homophonic translations do not necessarily take place with the
preservation of meaning---which often results in outrageous
mistranslations like the ones aforementioned---homophonic translations
sometimes may lead to correct translations. For example, imagine a
Spanish native with no previous familiarity with any other language
listening to French for the first time. This listener may encounter the
word /\textipa{"pOKt}/ (\emph{porte}), translation of door in French.
When comparing the phonemic transcription of the French to that of its
Spanish translation /\textipa{"pweR.ta}/ (\emph{puerta}), one may be led
to infer that these two words do not have much overlap. But one thing to
note is that the voiced uvular fricative consonant /\textipa{K}/ and the
open-mid back rounded vowel /\textipa{O}/, which do not exist in Spanish
as phonemic categories, can be perceived as allophonic variations the
native (Spanish) phonemes /\textipa{R}/ and /o/, as described by
assimilation models of phoneme perception
(\citeproc{ref-best1988examination}{Best et al., 1988},
\citeproc{ref-best2001discrimination}{2001}). Although this phonetic
mismatch between the non-native word and its translation can have a
noticeable toll on comprehension Cutler et al.
(\citeproc{ref-cutler2004patterns}{2004}), the Spanish listener may
still be able to activate the phonological representation of
/\textipa{"pweR.ta}/ (\emph{puerta}) from hearing the word-form
/\textipa{"pOKt}/ (\emph{porte}), ultimately allowing them to access the
correct semantic representation. Given enough phonological similarity
between the non-native word and its translation in the native language,
a naïve listener may succeed at word comprehension when listening to an
unfamiliar language. Such phonological similarity between translation
equivalents---known as \emph{cognateness}---is common across many
languages, and is often due to typological closeness and/or
socio-historical events involving the speakers of these languages (e.g.,
migration, social contact). For example, Romance languages such as
Spanish and Catalan share many form-similar translation equivalents
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and orthographic similarity of the former to the equivalent
in their native language.

While Soramimi (and more generally homophonic translations) are popular
and frequent phenomena, the psycholinguistic literature on the topic is
relatively scarce, and has mostly focused on the phonological variables
involved in the occurence of Soramimi. For instance, Otake
(\citeproc{ref-otake2007interlingual}{2007}) analysed 194 instances of
Soramimi, broadcasted between 1992 and 2007 by the TV show Soramimi
hour. These instances consisted of homophonic translations of English
song lyrics to Japanese words and phrases. The vast majority of the
reported instances were phrases (96\%) and a few, to single words (4\%).
The phonetic features of the presented English lyrics were preserved
with relatively high variability, with some Japanese resulting words or
phrases sharing high overlap with the original English strings of
sounds, and some sharing very little overlap. When analysing the few
instances of homophonic translations of single words, the author
identified three phonological processes that might explain how Japanese
listeners reconstructed the English input to generate the Japanese
words: insertion (e.g., \emph{cry} /\textipa{kraI}/ to \emph{kurai},
/\textipa{kuRai}/ {[}dark{]}), deletion (e.g., \emph{go},
/\textipa{goU}/ to \emph{go}, /go/ {[}go{]}, and alternation (e.g.,
\emph{low}, /\textipa{loU}/ to \emph{rou}, /\textipa{Roo}/ {[}wax{]}).
This suggests that the reported homophonic translations can be
explained, to some extent, as the result of the Japanese listeners
accommodating the strings of English sounds to the Japanese phoneme
inventory and Japanese phonotactics
(\citeproc{ref-dupoux1999epenthetic}{Dupoux et al., 1999};
\citeproc{ref-peperkamp2008perceptual}{Peperkamp et al., 2008}). Native
prosody also constrains the occurence of homophonic translations. For
instance, Kentner (\citeproc{ref-kentner2015rhythmic}{2015}) presented
in the auditory modality French-native and German-native participants
with English songs with lyrics. Homophonic translations followed
language-specific segmentation strategies. Most homophonic translations
reported by German-native participants resulted from inserting phrase
and word boundaries \emph{before} stressed syllables (following the more
frequent trochaic patter in German). On the contrary, most homophonic
translations reported by French-native participants resulted from
inserting phrase and word boundaries \emph{after} stressed syllables,
following the more frequent iambic pattern in French.

These studies have addressed the phonological aspects of homophonic
translation, but little attention has been paid to the dynamics of
lexical activation and competition that underlay this phenomenon. In the
present work, we investigated the interplay between phonological
similarity and phonological neighbourhood density in a non-native speech
perception task. In particular, we studied how the presence of
phonological neighbours in the native language lexicon affects the
extent to which listeners are able to activate the correct translation
of an auditorily presented word in thier native language. We used a
translation elicitation task in which participants listened to words
from an unfamiliar language, and then tried to guess their translation
in their native language. We will henceforth refer to the
auditory-presented words heard by participants on each trial as
\emph{presented words}, and the correct translation for the presented
words as \emph{target translations}. We explored listeners' reliance on
phonological similarity to succeed in the task by manipulating the
amount of phonological similarity between the presented words and their
target translations. Since participants were unfamiliar with the
presented language, they could only use phonological similarity between
the presented language and their native language to successfully
translate the words. We therefore predicted that participants'
performance should increase when the translation pairs are
phonologically more similar. We also predicted that there is a minimum
threshold of phonological similarity to be sufficient for translation.

@@

In Experiment 1, we collected data from two groups of British English
natives living in the United Kingdom. One group was presented with
Catalan words, the other with Spanish words. We examined the extent to
which participants were able to use the phonological similarity between
the presented word (in Catalan or Spanish) and its target translation to
provide accurate responses. In Experiment 2, we tested a group of
(European) Spanish natives in the same task, who were presented with
Catalan words. Catalan and Spanish are two Romance languages whose close
typological distance is reflected in the fact that they share many
cognates, where English is a Germanic language that shares considerably
fewer cognates with Catalan and Spanish. By testing participants
translating words from typologically close or distant languages, we
expected to widen the range of the phonological similarity scores of the
translation pairs involved in the experimental task, therefore allowing
us to explore potential cross-language differences in participant's
performance. One unexpected finding was that participants in Experiments
1 and 2 were surprisingly good at translating a subset of words which
had low phonological similarity with their correct translation. We were
concerned that this may be caused by some prior knowledge of specific
words by our participants, as some Spanish words are commonly seen in
media or product labels, making them familiar even to monolingual
speakers of English. In Experiment 3, we collected additional data from
a new group of British English natives. The design was closely modelled
after Experiment 1, except that after providing their response in each
trial, participants reported whether they had previous knowledge of the
presented word. In the final section of this article, we present
analyses on the joint data sets of Experiments 1 to 3.

@@

We collected data from 71 British English-native adults living in United
Kingdom (\emph{Mean} = 21.76 years, \emph{SD} = 2.15, Range = 18-26, 46
female). Data collection took place from June 04th, 2020 to June 25th,
2020. Participants were recruited via Prolific (5£ compensation) and
SONA (compensation in academic credits). Participants gave informed
consent before providing any data. The study was conducted in accordance
with ethical standards of the Declaration of Helsinki and the protocol
was by the University of Oxford Medical Sciences Inter-Divisional
Research Ethics Committee (IDREC) (R60939/RE007). Participants were
asked to complete the experiment using a laptop in a quiet place with
good internet connection.

@@

\caption{\label{tbl-participants}Participant details.}

@@

We created two lists of input words to be presented to participants in
the auditory modality: one in Catalan and one in Spanish. Words in the
Catalan list were 5.02 phonemes long on average (\emph{SD} = 1.49, Range
= 2-8), and the orthographic forms of their English translations (which
participants had to type) were 5.12 characters long on average
(\emph{SD} = 1.56, Range = 3-9). Words in the Spanish list were 5.52
phonemes long on average (\emph{SD} = 1.47, Range = 3-9), and the
orthographic form of their English translations were 5.29 characters
long on average (\emph{SD} = 1.77, Range = 3-12).

In each trial, participants listened to one audio file, wich contained a
single word. The audio files were the same ones used in child
experiments conducted in the Laboratori de Recerca en Infància of
Universitat Pompeu Fabra (Barcelona, Spain). These audio files were
recorded by a proficient Catalan-Spanish female bilingual from the
Metropolitan Area of Barcelona in a child-directed manner. Catalan and
Spanish words were recorded at 44,100 Hz in separate files in the same
session, and then de-noised using Audacity and normalised at peak
intensity using Praat (\citeproc{ref-broersma2021praat}{Broersma \&
Weenink, 2021}). The average duration of the Catalan audio files was
1.24 seconds (\emph{SD} = 0.19, Range = 0.8-1.58). The average duration
of the Spanish audio files was 1.16 seconds (\emph{SD} = 0.15, Range =
0.78-1.53).

For each word in the Catalan and Spanish lists pair, we defined three
predictors of interest: the lexical frequency of the correct English
translation (\emph{Frequency}), the phonological similarity between the
presented word (in Catalan or Spanish) and their correct English
translation (\emph{Similarity}), and the presented word's number of
cross-language phonological neighbours (\emph{CLPN}) in English.
\emph{Frequency} was included as a nuance predictor, under the
hypothesis that---keeping other predictors constant---participants would
translate higher-frequency words more accurately and faster than
lower-frequency words. Lexical frequencies of correct translations were
extracted from SUBTLEX-UK (\citeproc{ref-van2014subtlex}{Van Heuven et
al., 2014}), and transformed to Zipf scores. Words in the stimuli list
without a lexical frequency value were excluded from data analysis (2 in
the Catalan list, 5 in the Spanish list).

We calculated \emph{Similarity}, our main predictor of interest, by
computing the Levenshtein similarity between the X-SAMPA transcriptions
of each pair of translations using the \texttt{stringdist} R package
(\citeproc{ref-van2014stringdist}{van der Loo, 2014}). The Levenshtein
distance computes the edit distance between two character strings---in
this case, two phonological transcriptions---by counting the number of
additions, deletions, and substitutions necessary to make both strings
identical (\citeproc{ref-levenshtein1966binary}{Levenshtein, 1966}). We
divided this edit distance by the number of characters included in the
longest X-SAMPA transcription of the translation pair. This results in a
proportion score, in which values closer to zero correspond to lower
Levenshtein distances between phonological transcriptions (i.e., higher
similarity), and values closer to 1 correspond to higher Levenshtein
distances (i.e., lower similarity). This transformation accounts for the
fact that the Levenshtein distance tends to increase with the length of
the transcriptions. For interpretability, we subtracted this proportion
from one, so that values closer to one correspond to higher similarity
between phonological transcriptions, and values closer to zero
correspond to lower similarity between phonological transcriptions. For
example, the \emph{table} (\texttt{teIb@l})-\emph{mesa} (\texttt{mesa})
translation pair had a 17\% similarity, while the \emph{train}
(\texttt{trEIn})-\emph{tren} (\texttt{tREn}) translation pair had a 60\%
similarity. Table~\ref{tbl-stimuli} summarises the lexical frequency,
phonological neighbourhood density and phonological overlap of the words
included in the Catalan and the Spanish lists.
